\ \\ \LARGE\bf Generating Lambda Term Individuals in Typed Genetic Programming Using Forgetful A*

### 2014-01-20 tex/cec14/article.tex
cecky..

\begin{abstract}
In this paper, a generalization of genetic programming 
for simply typed lambda calculus is presented. We propose three population 
initialization methods depending on different search strategies. 
The first strategy corresponds to standard ramped half-and-half initialization, 
the second one corresponds to exhaustive systematic search, and the third one 
represents a novel geometric strategy. Three well-known benchmark experiments
support that the geometric strategy outperforms
the standard generating method in success rate, time consumption and average individual size. 
\end{abstract}
% no key words

@@

In \cite{yu01} Yu presents a GP system utilizing
polymorphic higher-order functions\footnote{Higher-order 
function is a function taking another function as 
input parameter.} and lambda abstractions.
Important point of interest in this work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion},
i.e. recursion without explicit recursive calls. 
The terminal set for constructing lambda abstraction subtrees 
is limited to use only constants and variables of that particular
lambda abstraction, i.e., outer variables are not allowed to be used
as terminals in this work. This is significant difference from our approach 
since we permit all well-typed normalized \lterms. From this difference also
comes different crossover operation. We focus more on term generating process; 
their term generation is performed in a similar way as the standard one, 
whereas our term generation also tries to utilize techniques of systematic enumeration.

In \cite{kes} Briggs and O’Neill present technique 
utilizing typed GP with combinators.
The difference between approach presented in this work
and our approach is that in this work terms are generated
straight from \textit{library} of combinators and no lambda abstractions
are used. They are using more general polymorphic type system then us
-- the Hindley–Milner type system. They also discuss the 
properties of exhaustive enumeration of terms and compare it with GP search.  
They also present interesting concept of \textit{Generalized
genetic operator} based on term generation.

In \cite{binard2008genetic} by Binard and Felty even 
stronger type system (\textit{System F}) is used.  
But with increasing power of the type system comes increasing difficulty of term generation.
For this reason evolution in this work takes interesting and nonstandard shape 
(fitness is associated with \textit{genes} which are evolved together with \textit{species}
which together participate in creation of individuals).
This differs from our approach, which tries to be generalization of
the standard GP\cite{koza92}.

In contrast with above mentioned works our approach uses very simple type system 
(simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do
so we use technique based on \textit{inhabitation machines} 
described by Barendregt in~\cite{barendregt10}.

@@

@@

\begin{figure}[h!]
  \caption{Graphs for Simple symbolic regression.}
  \centering
    \includegraphics[scale=0.15]{imgs/cec/SSR.PNG}
\end{figure}

@@

Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
Minimal $I(M,i,z)$ was in generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 76.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
Minimal $I(M,i,z)$ was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 47.
The experiment took 26 minutes.
Figure~1 shows the progress of those metrics throughout generations.
So for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful then 
ramped half-and-half strategy in all observed metrics.

@@

\begin{figure}[h!]
  \caption{Graphs for Artificial ant problem.}
  \centering
    \includegraphics[scale=0.15]{imgs/cec/ANT.PNG}
\end{figure}

@@

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 0.7999030182 & 0.8174067512 \\
SD	 & 0.1521369287	& 0.1826109800 \\
SEM	 & 0.0215154108	& 0.0258250925 \\
N	 & 50 & 50  \\
\hline
t-value & 0.5207 & \\
p-value & 0.6038 & \\
\hline
\end{tabular}
\caption{Statistical analysis of Simple symbolic regression.
By conventional criteria, this difference is considered to be not statistically significant.}
\end{table}

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 66.20 & 79.78 \\
SD	 & 12.33 & 8.77  \\
SEM	 & 1.74	 & 1.24  \\
N	 & 50    & 50    \\
\hline
t-value &  6.3479           & \\
p-value &  less than 0.0001 &  \\
\hline
\end{tabular}
\caption{Statistical analysis of Artificial ant problem.
 By conventional criteria, this difference is considered to be extremely statistically significant. }
\end{table}

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 0.895000	& 0.930000 \\
SD	 & 0.075170	& 0.078490 \\
SEM	 & 0.010631	& 0.011100 \\
N	 & 50    & 50    \\
\hline
t-value &  2.2772 & \\
p-value &  0.0250 &  \\
\hline
\end{tabular}
\caption{Statistical analysis of Even parity problem. 
By conventional criteria, this difference is considered to be statistically significant.}
\end{table}

Standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
Minimal $I(M,i,z)$ was in generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was circa 400.
The experiment took 265 minutes.
Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
Minimal $I(M,i,z)$ was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was circa 90.
The experiment took 107 minutes.
This is a big improvement in all watched factors.
Figure~2 shows it in greater detail.

@@

Previous two problems were instances of standard GP, i.e. their building symbols
obeyed the closure requirement. This third experiment will break the closure requirement, thus ideal for testing typed GP techniques. 
The even-parity function is a boolean function taking as inputs $N$
boolean values and returning \textit{True} if an even number of inputs 
are \textit{True}. For odd number it returns \textit{False}.
This problem has been used by many researchers
as a benchmark for GP. We compare our results with that obtained by Yu in \cite{yu01},
where is presented approach to evolve recursive and modular programs
by use of higher-order functions and $\lambda$-abstractions.
We use very similar set of \textit{building symbols} as in \cite{yu01}. 
\begin{align*}
\tau = [Boo&l] \ar Bool\\
\Gamma = \{
  and   &: Bool \ar Bool \ar Bool                              ,\\
  or    &: Bool \ar Bool \ar Bool                              ,\\
  nand  &: Bool \ar Bool \ar Bool                              ,\\
  nor   &: Bool \ar Bool \ar Bool                              ,\\
  foldr &: (Bool \ar Bool \ar Bool) \\
        &~~~\ar Bool \ar [Bool] \ar Bool ,\\
  head' &: [Bool] \ar Bool                                   ,\\
  tail' &: [Bool] \ar [Bool]                              \}
\end{align*}
\begin{figure}[h!]
  \caption{Graphs for Even parity problem.}
  \centering
    \includegraphics[scale=0.15]{imgs/cec/EP.PNG}
\end{figure}

@@

Standard \textit{ramped half-and-half strategy} scored 9/50 (18\%) success rate. 
Minimal $I(M,i,z)$ was in generation 17 with 333,000 individuals to be processed.
The average individual size for generation 50 was 39.
The experiment took 28 minutes.
Our \textit{geometric strategy} scored 32/50 (64\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was circa 100.
The experiment took 33 minutes.
Figure~3 shows it in greater detail.
Once again this is a big improvement against standard method. 
One may get the impression that geometric strategy suffers from bloat here,
but considering that it performs significantly better and
that term size around 100 is similar as for the previous two problems, 
it seems to us that ramped half-and-half maybe has difficulties with 
constructing more complex terms, and therefore is "staying small".
Another remarkable observation is that 4 times out of 50 (8\% of all runs) 
the 100\% correct solution was present in the generation 0. This is pretty good
result for uninformed search. However, our results are less successful 
then those presented in \cite{yu01} (with slightly different set of 
building symbols); they scored 40/50 (80\%) success rate and $I(M,i,z)$ of 
17,500 in generation 4. But by observation of typical solution which is often
some modification of \texttt{foldr xor False inputList} one sees that important task
for GP here is to create \texttt{xor} function. This is advantage for the more specialized term representation used in \cite{yu01} 
(restricting use of outer variables). 
The difference in crossover operators may also be involved in the difference, 
but such analysis is beyond the scope of this paper. Finally, our result at least outperforms other results mentioned in \cite{yu01}: \textit{Generic genetic programming} 
scored 17/60 (28\%); min $I(M,i,z) = 220,000$. \textit{GP with ADFs} 
scored 10/29 (34\%); min $I(M,i,z) = 1,440,000$.

@@

>>>>

\begin{abstract}
Tree based genetic programming traditionally uses simple S-expressions to represent programs, however more expressive representations can exhibit better results and are well suited for typed genetic programming. One of such representations is lambda calculus.
In this paper, a generalization of genetic programming 
for simply typed lambda calculus is presented. We mainly focus on
population initialization methods. 
Three such initialization methods are proposed 
depending on different search strategies. 
The first strategy corresponds to standard ramped half-and-half initialization, 
the second one corresponds to exhaustive systematic search, and the third one 
represents a novel geometric strategy. 
Three well-known benchmark experiments (simple symbolic regression,
artificial ant problem and even parity problem)
support that the geometric strategy outperforms
the standard generating method in success rate, best fitness value,
time consumption and average individual size. 
Despite the fact that the presented system is designed for typed genetic 
programming, when given standard genetic programming problem satisfying
standard closure requirement it produces simple S-expressions and
the novel geometric strategy outperforms the standard system. 
This claim is supported by that first two experiments both satisfying 
the closure requirement.
The fact that a search strategy can be easily described by a rather simple filtration algorithm opens opportunity to easy exploration of more 
population initialization methods, possibly in a kind of meta-evolution way.
\end{abstract}
% no key words

@@

Yu presents a GP system utilizing
polymorphic higher-order functions\footnote{Higher-order 
function is a function taking another function as 
input parameter.} and lambda abstractions  \cite{yu01}.
Important point of interest in this work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion},
i.e. recursion without explicit recursive calls. 
The terminal set for constructing lambda abstraction subtrees 
is limited to use only constants and variables of that particular
lambda abstraction, i.e., outer variables are not allowed to be used
as terminals in this work. This is significant difference from our approach 
since we permit all well-typed normalized \lterms. From this difference also
comes different crossover operation. We focus more on term generating process; 
their term generation is performed in a similar way as the standard one, 
whereas our term generation also tries to utilize techniques of systematic enumeration.

Briggs and O’Neill present technique 
utilizing typed GP with combinators \cite{kes}.
The difference between approach presented in this work
and our approach is that in this work terms are generated
straight from \textit{library} of combinators and no lambda abstractions
are used. They are using more general polymorphic type system then us
-- the Hindley–Milner type system. They also discuss the 
properties of exhaustive enumeration of terms and compare it with GP search.  
They also present interesting concept of \textit{Generalized
genetic operator} based on term generation.

Binard and Felty use even 
stronger type system (\textit{System F}) \cite{binard2008genetic}.  
But with increasing power of the type system comes increasing difficulty of term generation.
For this reason evolution in this work takes interesting and nonstandard shape 
(fitness is associated with \textit{genes} which are evolved together with \textit{species}
which together participate in creation of individuals).
This differs from our approach, which tries to be generalization of
the standard GP\cite{koza92}.

In contrast with above mentioned works our approach uses very simple type system 
(simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do
so we use technique based on \textit{inhabitation machines} 
described by Barendregt \cite{barendregt10}.

@@

\red{Přestože je naše generujicí metoda navržena obecně pro typovanou verzi,
tak pracuje dobře i pro problémy standardního GP splňující closure podmínku.
Pro taklovéto problémy strategie generuje standardní S-výrazy. Proto jsou
první dva ze tří experimentů právě takového charakteru. Dalším důvodem pro volbu 
takovýchto problémů je fakt, že výsledky pokusů výrazně ovlivňuje použitý křížící 
operátor. Návrh obecného křížícího operatoru pro GP v symply typed lc je otevřený problém. Protože křížící operátor použitý v našem systému je poměrně komplikovaný
a tím pádem beyond the scope of this paper, ale pro problémy splnující
clousure podmínku se chová přesně stejně jako standardní křížící operator,
tak se to zdá jako další důvod proč 2/3 experimentů udělat na closed problémech.}

@@

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/SSR.PNG}
  \caption{Graphs for Simple symbolic regression.}
\end{figure}

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 0.7999030182 & 0.8174067512 \\
SD	 & 0.1521369287	& 0.1826109800 \\
SEM	 & 0.0215154108	& 0.0258250925 \\
N	 & 50 & 50  \\
\hline
t-value & 0.5207 & $\alpha = 0.05$\\
p-value & 0.6038 & \textbf{Not statistically significant.}\\
\hline
\end{tabular}
\caption{Statistical analysis - Simple symbolic regression.}
\end{table}

~\\

Figure~1 shows results of this experiment. 
Table~I summarizes statistical analysis
of the fitness values of the best individuals of the run.  
By conventional criteria, this difference is considered to be not statistically significant.
Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
Minimal $I(M,i,z)$ was in generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 68.8.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
Minimal $I(M,i,z)$ was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 44.1.
The experiment took 26 minutes.
So for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful then 
ramped half-and-half strategy in these observed metrics.

@@

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/ANT.PNG}
  \caption{Graphs for Artificial ant problem.}
\end{figure}

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 66.20 & 79.78 \\
SD	 & 12.33 & 8.77  \\
SEM	 & 1.74	 & 1.24  \\
N	 & 50    & 50    \\
\hline
t-value &  6.3479           & $\alpha = 0.05$\\
p-value &  \textless 0.0001 &  \textbf{Extremely statistically significant.}\\
\hline
\end{tabular}
\caption{Statistical analysis - Artificial ant problem.}
\end{table}

Figure~2 shows results of this experiment. 
Table~II summarizes statistical analysis
of the fitness values of the best individuals of the run.  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be extremely statistically significant.

Standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
Minimal $I(M,i,z)$ was in generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was 279.5.
The experiment took 265 minutes.

Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
Minimal $I(M,i,z)$ was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was 94.1.
The experiment took 107 minutes.

This is a big improvement in all watched factors.

@@

Previous two problems were instances of standard GP, i.e. their building 
symbols obeyed the closure requirement. This third experiment will 
break the closure requirement, thus ideal for testing typed GP techniques. 
The even-parity function is a boolean function taking as inputs $N$
boolean values and returning \textit{True} if an even number of inputs 
are \textit{True}. For odd number it returns \textit{False}.
This problem has been used by many researchers
as a benchmark for GP. We compare our results with 
that obtained by Yu in \cite{yu01},
where is presented approach to evolve recursive and modular programs
by use of higher-order functions and $\lambda$-abstractions.
We use very similar set of \textit{building symbols} as in \cite{yu01}. 
\begin{align*}
\tau = [Boo&l] \ar Bool\\
\Gamma = \{
  and   &: Bool \ar Bool \ar Bool                              ,\\
  or    &: Bool \ar Bool \ar Bool                              ,\\
  nand  &: Bool \ar Bool \ar Bool                              ,\\
  nor   &: Bool \ar Bool \ar Bool                              ,\\
  foldr &: (Bool \ar Bool \ar Bool) \\
        &~~~\ar Bool \ar [Bool] \ar Bool ,\\
  head' &: [Bool] \ar Bool                                   ,\\
  tail' &: [Bool] \ar [Bool]                              \}
\end{align*}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/EP.PNG}
  \caption{Graphs for Even parity problem.}
\end{figure}

\begin{table}[t]
\centering
\begin{tabular}{|l|cc|}
\hline
& Ramped half-and-half & Geometric \\
\hline
Mean & 0.895000	& 0.930000 \\
SD	 & 0.075170	& 0.078490 \\
SEM	 & 0.010631	& 0.011100 \\
N	 & 50    & 50    \\
\hline
t-value &  2.2772 & $\alpha = 0.05$\\
p-value &  0.0250 &  \textbf{Statistically significant.}\\
\hline
\end{tabular}
\caption{Statistical analysis - Even parity problem.}
\end{table}

Figure~3 shows results of this experiment. 
Table~III summarizes statistical analysis
of the fitness values of the best individuals of the run.  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be 
statistically significant.

Standard \textit{ramped half-and-half strategy} scored 9/50 (18\%) success rate. 
Minimal $I(M,i,z)$ was in generation 17 with 333,000 individuals to be processed.
The average individual size for generation 50 was 47.7.
The experiment took 28 minutes.
Our \textit{geometric strategy} scored 32/50 (64\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was 74.3.
The experiment took 33 minutes.

Once again this is a big improvement against standard method.

One may get the impression that geometric strategy suffers from bloat here,
but considering that it performs significantly better and
that term size around 100 is similar as for the previous two problems, 
it seems to us that ramped half-and-half maybe has difficulties with 
constructing more complex terms, and therefore is "staying small".
Another remarkable observation is that 4 times out of 50 (8\% of all runs) 
the 100\% correct solution was present in the generation 0. This is pretty good
result for uninformed search. However, our results are less successful 
then those presented in \cite{yu01} (with slightly different set of 
building symbols); they scored 40/50 (80\%) success rate and $I(M,i,z)$ of 
17,500 in generation 4. But by observation of typical solution which is often
some modification of \texttt{foldr xor False inputList} one sees that important 
task for GP here is to create \texttt{xor} function. This is advantage for the 
more specialized term representation used in \cite{yu01} 
(restricting use of outer variables). 
The difference in crossover operators may also be involved in the difference, 
but such analysis is beyond the scope of this paper. Finally, our result at least outperforms other results mentioned in \cite{yu01}: \textit{Generic genetic programming} 
scored 17/60 (28\%); min $I(M,i,z) = 220,000$. \textit{GP with ADFs} 
scored 10/29 (34\%); min $I(M,i,z) = 1,440,000$.

@@

\red{Ještě nějak rozšířit.}
